Peer Assignment 1
===================

This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps
taken in 5 minute intervals each day.

*The first task is to retrieve, and pre-process the data

```{r, echo=TRUE}
activity <- read.csv("activity.csv")
```

We must then create a date.time column that will combine the date and interval columns

```{r,echo=TRUE}
time <- formatC(activity$interval/100, 2, format = "f")
activity$date.time <- as.POSIXct(paste(activity$date, time), format = "%Y-%m-%d %H.%M", 
    tz = "GMT")
```

We then create a time column to make it easier to to analyse the mean for different time of day.
All the times are thus converted into today. 

```{r,echo=TRUE}
activity$time <- format(activity$date.time, format = "%H:%M:%S")
activity$time <- as.POSIXct(activity$time, format = "%H:%M:%S")
```

*What is mean total number of steps taken per day?*

**What is the mean number of steps for each day?

```{r,echo=TRUE}
Total.Steps <- tapply(activity$steps, activity$date, sum, na.rm = TRUE)
```

**Mean total steps per day

```{r,echo=TRUE}
mean(Total.Steps)
```

**Median total steps per day
```{r,echo=TRUE}
median(total.steps)
```

**Using a histogram we examine the distribution of total number of steps per day
```{r,echo=TRUE}
library(ggplot2)
qplot(total.steps, xlab = "Total steps", ylab = "Frequency")
## stat_bin: binwidth defaulted to range/30. Use 'binwidth = x' to adjust this.
```

*What is the average daily activity pattern*

**The mean steps for each five minute interval is calculated, and then put in a data frame

```{r,echo=TRUE}
mean.steps <- tapply(activity$steps, activity$time, mean, na.rm = TRUE)
daily.pattern <- data.frame(time = as.POSIXct(names(mean.steps)), mean.steps = mean.steps)
```

**A time series plot is then calculated for the mean steps

```{r,echo=TRUE}
library(scales)
ggplot(daily.pattern, aes(time, mean.steps)) + geom_line() + xlab("Time of day") + 
    ylab("Mean number of steps") + scale_x_datetime(labels = date_format(format = "%H:%M"))
```

**Which five minute has the highest mean number of steps?

```{r,echo=TRUE}
most <- which.max(daily.pattern$mean.steps)
format(daily.pattern[most, "time"], format = "%H:%M")
```

*Imputing Missing Data*

** The number of intervals with missing steps counts("NA's") is identified

```{r,echo=TRUE}
summary(activity$steps)
```

**To fill in the missing values, we can use mean steps for a five-minute interval for the entire dataset

```{r,echo=TRUE}
library(Hmisc)
activity.imputed <- activity
activity.imputed$steps <- with(activity.imputed, impute(steps, mean))
```

**The mean and median steps for each day between the original data set and 
the imputed data set are then computed

```{r,echo=TRUE}
total.steps.imputed <- tapply(activity.imputed$steps, activity.imputed$date, 
    sum)
mean(total.steps)
```

```{r,echo=TRUE}
mean(total.steps.imputed)
```

```{r,echo=TRUE}
median(total.steps)
```

```{r,echo=TRUE}
median(total.steps.imputed)
```

**A histogram of the imported dataset is then created

```{r,echo=TRUE}
qplot(total.steps.imputed, xlab = "Total steps", ylab = "Frequency")
## stat_bin: binwidth defaulted to range/30. Use 'binwidth = x' to adjust this
```

**Imputing the missing data has increased the average number of steps

*Are there differences in activity patterns between weekdays and weekends?*

**A factor column is added for whether a day is a weekday or a weekend

```{r,echo=TRUE}
day.type <- function(date) {
    if (weekdays(date) %in% c("Saturday", "Sunday")) {
        return("weekend")
    } else {
        return("weekday")
    }
}

day.types <- sapply(activity.imputed$date.time, day.type)
activity.imputed$day.type <- as.factor(day.types)
```

**Create a dataframe that holds the mean steps for weekdays and weekends

```{r,echo=TRUE}
mean.steps <- tapply(activity.imputed$steps, interaction(activity.imputed$time, 
    activity.imputed$day.type), mean, na.rm = TRUE)
day.type.pattern <- data.frame(time = as.POSIXct(names(mean.steps)), mean.steps = mean.steps, 
    day.type = as.factor(c(rep("weekday", 288), rep("weekend", 288))))
```

**Compare the patterns between weekdays and weekends

```{r,echo=TRUE}
ggplot(day.type.pattern, aes(time, mean.steps)) + geom_line() + xlab("Time of day") + 
    ylab("Mean number of steps") + scale_x_datetime(labels = date_format(format = "%H:%M")) + 
    facet_grid(. ~ day.type)
```





























